Characterizing the response of subarctic European vegetation to key environmental variables for land surface modeling of ozone pollution and risk assessment

### 2021-10-14 VegOzone.tex
Conclude suggestios by Ane and Aud

The climatology for global irradiance ($Q_0$, not shown) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). We select the highest $Q_0$ for each day (noon) and average each over the whole observation period. Thus, we find the highest averaged daily maximum ($\left<Q_0^\mathrm{max}\right>$) which amounts to $800\,\unit{W\,m^{-2}}$. Similarly, we diurnal mean $Q_0$ (including night) and find the maximum of the averaged daily mean ($\left<Q_0^\mathrm{mean}\right>$) $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). By selecting the lowest daily $Q_0$ (including night), we compute the maximum of the averaged daily minimum ($\left<Q_0^\mathrm{min}\right>$) about $200\,\unit{W\,m^{-2}}$. This indicates that a considerable amount of light is also available during night.

%Ozone
The derived ozone climatology (\chem{\left<[O_3]\right>}) (1986--1996) is shown in Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. A spline was fitted through the data to guide the eye. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display the expected, pronounced seasonal cycle. The spring peak occurs on average on day of the year (\unit{doy}) $100$ and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $28\,\unit{ppb}$. The decline in \chem{\left<[O_3]\right>} coincides with the average begin of \chem{CO_2} uptake by coniferous trees \citep{TB:Kolari2007, TP:Wallin2013}. In July--September, ozone is occasionally almost completely depleted. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air.

@@

We define two different hypothetical temperature acclimations: subarctic and cold. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. This is accomplished by moving $T_\mathrm{opt}$ towards cooler temperatures while keeping the other parameters fixed to MM values. In the same way, \emph{subarctic} is constructed to represent a species which is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift $T\mathrm{min}$ and $T\mathrm{max}$ to colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. In other words, we aim to increase the overlapping area between $f_\mathrm{temp}$ and the temperature PDF during the GS. The most extreme of the two acclimation types is \emph{subarctic}.
In Fig.~\ref{fig:f_temp_grassland}a, we show the resulting $f_\mathrm{temp}$ for perennial grassland together with temperature PDFs deduced from the five months GS in two climatological periods. The periods 1990s and 2000s have been chosen to indicate that the vegetation has been subject to climate change. Thus, the predominant acclimation at present is not known.

@@

Regarding the acclimation of $f_\mathrm{light}$, we presume that the opening extent of stomata at low light conditions differs in subarctic species compared to species in less extreme climates. We, therefore, look at the PPFD threshold for a $50\,\unit{\%}$ opening of plants' stomata. We derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) analytically for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. By inverting Eq.~(\ref{eq:inverse_function}) for $\alpha$, we find $\alpha'(\gamma')$. The resulting functions for perennial grassland (PPFD0.8, PPFD1.2) are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b and \ref{fig:f_temp_birch}b. All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

Another important factor is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product \citep{MODIS_PSN} for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start}$ as \unit{doy}~122 for 2018 and \unit{doy}~106 in 2019. $A_\text{end}$ amounts to \unit{doy}~261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 is, hence, one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find \unit{doy}~129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR). For comparison, the generic dates for central Europe ($A_\mathrm{start} = 100$ and $A_\mathrm{end} = 307$) indicating a substantially longer growing season. Due to a lack of quantitative field observation at Svanhovd, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}.

@@

Soil texture influences water availability and thus $\mathrm{POD_y}$ calculations in $\mathrm{DO_3SE}$ \citep{ACP:Bueker2012}. The soil at Svanhovd is characterized in an ICP Forest plot survey as gleyic cambisols according to the Food and Agriculture Organization of the United Nations (FAO) soil classification system and as eluviated dystric brunisol in the ICP Forest soil classification system (V. Timmerman, personal communication, May 2021) Within the $\mathrm{DO_3SE}$ soil parameterization, the sandy loam texture describes the upper $60\,\unit{cm}$, where the plant roots are found, of this soil type best.

The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. A sample of downy birch leaves collected at Svanhovd had an average length of $(3.0\pm 0.5)\,\unit{cm}$ which is smaller than in the MM parameterization. \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. One Scots pine forest at Svanhovd has been studied as a part of the ICP Forest mapping. Heights were measured in 2004, when the stand was 90 years old. Mean tree height was $10.1\,\unit{m}$ and maximum height was $16.2\,\unit{m}$ (V. Timmerman, personal communication, 2021 May 10).

@@

The magnitude of these effects varies between PFTs as well as years, but the predicted ozone uptake for the bespoke parameterization is always larger than for the MM parameterizations and of the same order of magnitude as the variability between the years studied here. Our bespoke temperature parameterizations suggest considerable underestimations of ozone risk for subarctic species when relying on the generic MM parameterizations.

@@

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). The MM Boreal Norway spruce arameterization was already well acclimated to prevailing temperatures and therefore differs least from our bespoke parameterizations. Taking the bespoke GS and canopy height into account, we find a higher ozone uptake in 2019 than in 2018. The overall difference deduced from bespoke GS for subarctic and cold parameterizations with respect to MM range between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ and amounts to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

@@

@@

We estimated the biomass reductions in accordance to the CLs \citep{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature parameterization (\emph{subarctic}) for each bespoke PFT. The estimates are listed in Table~\ref{tab:biomass_reduction}. The reported range corresponds to the maximum deviation defined by the $\pm 20\,\unit{\%}$ variation of $f_\mathrm{light}$ with SWP=on. Total biomass of deciduous trees and perennial grassland was substentially reduced in comparison with the default MM parameterizations in both years. The CL for perennial grassland is higher than for the other species. From an agro-economical point of view in our focus area, a loss of biomass of more than $10\,\unit{\%}$ can be more severe compared to more productive areas in central Europe and may need adjustment.

@@

From the difference between 2018 and 2019, we infer a substantial interannual variability. The interannual variability appears to be large and is of the same order of magnitude as our reported deviations compared to MM due to temperature and light acclimations. This means that the choice of temperature acclimation has the potential to either revoke or amplify the risk of ozone damage for vegetation.

@@

 The $\mathrm{DO_3SE}$ modeling results are in contrast to our observation of more visible damage in 2018 compared to 2019. However, visible damage caused by peak ozone uptake and a dependence of ozone sensibility on phenology are not accounted for in $\mathrm{POD_y}$. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland have been found to be more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed ozone damage to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009} but not accounted for in the models. This may also explain some contradictory results regarding ozone sensitivity of natural vegetation in northern Europe. \citet{FS:Subramanian2014} reported a higher, modeled reduction of net primary production in conifers ($4.3-15.5\,\unit{\%}$) than birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, while \citet{Amb:Girgzdiene2009} observed more visible damage on deciduous trees than on Norway spruce in Lithuania. But visible damage and biomass reduction may not occur at the same rate and hence may not be interchangeable.

In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures leading to a higher efficiency in photosynthesis at cool temperatures than for central European species. To this end, we computed a PDF from observed temperature at Svanhovd (May--September, 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. Remark that these parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.

@@

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to a compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change and its predictions strongly depend on both the range of existing acclimations as well as the actual type acclimation.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on species in less extreme climates. In particular, CLs for deciduous and coniferous trees were breached in all years and forall parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constrains on productivity leading to more severe economical consequences.

@@

\authorcontribution{SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV contributed to framing this research article, has performed the on-site observation of vegetation damage induced by ozone, and provided advice on plant physiological processes. LE has contributed to framing of the this article, $\mathrm{DO_3SE}$ modeling, and development of the PDF-based temperature acclimation assessment. CO has collected PFT parameters from the literature, performed all $\mathrm{DO_3SE}$ simulations and validation. AE contributed with her experience regarding subarctic vegetation in Finnmark. FS contributed with an assessment of the 2018 meteorological conditions. TB gave guidance in a broader research sense. All authors contributed to discussions, writing, and conclusions.} %% this section is mandatory

@@

\begin{acknowledgements}
  We thank the NIBIO Environment Centre Svanhovd for their support.
  Bj{\o}rg Rognerud for processing SeNorge.no data with respect for begin of growing season.
  Tore Flatlandsmo Berglen for hourly pressure data from Svanhovd.
  This work was supported by the Research Council of Norway (NRC) grand No. 268073.
\end{acknowledgements}

>>>>

The climatology for global irradiance ($Q_0$, not shown) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). We deduce the corresponding highest $Q_0$ to $800\,\unit{W\,m^{-2}}$. The maximum of the lowest $Q_0$ (including night) amounts to $200\,\unit{W\,m^{-2}}$ indicating that a considerable amount of light is available under midnight sun conditions.

%Ozone
The daily mean ozone climatology (\chem{\left<[O_3]\right>}) (1986--1996) is shown in Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. A spline was fitted through the data to guide the eye. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display the expected, pronounced seasonal cycle. The spring peak occurs on average on day of the year (\unit{doy}) $100$ and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $28\,\unit{ppb}$. The decline in \chem{\left<[O_3]\right>} coincides with the average beginning of \chem{CO_2} uptake by coniferous trees \citep{TB:Kolari2007, TP:Wallin2013}. In July--September (\unit{doy}~182--273), ozone is occasionally almost completely depleted. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air.

@@

We define two different hypothetical temperature acclimations: subarctic and cold. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. This is accomplished by moving $T_\mathrm{opt}$ towards cooler temperatures while keeping the other parameters fixed to MM values. In the same way, \emph{subarctic} is constructed to represent a species which is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift $T\mathrm{min}$ and $T\mathrm{max}$ to colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. In other words, we aim to increase the overlapping area between $f_\mathrm{temp}$ and the temperature probability density function (PDF) during the GS. The most extreme of the two acclimation types is \emph{subarctic}.
In Fig.~\ref{fig:f_temp_grassland}a, we show the resulting $f_\mathrm{temp}$ for perennial grassland together with temperature PDFs deduced from the five months GS in two climatological periods. The periods 1990s and 2000s have been chosen to indicate that the vegetation has been subject to climate change. Thus, the predominant acclimation at present is not known.

@@

Regarding the acclimation of $f_\mathrm{light}$, we presume that the opening extent of stomata at low light conditions differs in subarctic species compared to species in less extreme climates. We, therefore, adjust the light sensitivity of stomata such that the PPFD value needed to reach $50\,\unit{\%}$ opening is varied $\pm 20\,\unit{\%}$. For this we derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) analytically for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. By inverting Eq.~(\ref{eq:inverse_function}) for $\alpha$, we find $\alpha'(\gamma')$. The resulting functions for perennial grassland (PPFD0.8, PPFD1.2) are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b and \ref{fig:f_temp_birch}b. All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

Another important factor is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. For coniferous trees, we use a MODIS (Aqua/Terra) retrieved net photosynthesis product \citep{MODIS_PSN} for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start}$ as \unit{doy}~122 for 2018 and \unit{doy}~106 in 2019. $A_\text{end}$ amounts to \unit{doy}~261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 is, hence, one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find \unit{doy}~129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR). For comparison, the generic dates for central Europe ($A_\mathrm{start} = 100$ and $A_\mathrm{end} = 307$) indicating a substantially longer growing season. Due to a lack of quantitative field observation at Svanhovd, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}.

@@

Soil texture influences water availability and thus $\mathrm{POD_y}$ calculations in $\mathrm{DO_3SE}$ \citep{ACP:Bueker2012}. The soil at Svanhovd is characterized in an ICP Forest plot survey as gleyic cambisols according to the Food and Agriculture Organization of the United Nations (FAO) soil classification system and as eluviated dystric brunisol in the ICP Forest soil classification system (V. Timmerman, personal communication, June 2020) Within the $\mathrm{DO_3SE}$ soil parameterization, the sandy loam texture describes the upper $60\,\unit{cm}$, where the plant roots are found, of this soil type best.

The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. A sample of downy birch leaves collected at Svanhovd had an average length of $(3.0\pm 0.5)\,\unit{cm}$ which is smaller than in the MM parameterization. \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. One Scots pine forest at Svanhovd has been studied as a part of the ICP Forest mapping. Heights were measured in 2004, when the stand was 90 years old. Mean tree height was $10.1\,\unit{m}$ and maximum height was $16.2\,\unit{m}$ (V. Timmerman, personal communication, June 2020). We used $13.5\,\unit{m}$ height for both deciduous and coniferous trees.

@@

The magnitude of these effects varies between PFTs as well as years, but the predicted ozone uptake for the bespoke temperature parameterization is always larger than for the MM parameterizations and of the same order of magnitude as the variability between the years studied here. Our bespoke temperature parameterizations suggest considerable underestimations of ozone risk for subarctic species when relying on the generic MM parameterizations.

@@

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). The MM Boreal Norway spruce parameterization was already well acclimated to the prevailing temperatures and therefore differs the least from our bespoke parameterizations. Taking the bespoke GS and canopy height into account, we find a higher ozone uptake in 2019 than in 2018. The overall difference deduced from bespoke GS for subarctic and cold parameterizations with respect to MM range between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ and amounts to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

@@

From the difference between 2018 and 2019, we infer a substantial interannual variability. The maximum deviation from the generic default parameterizations due to our bespoke temperature and light response functions is of the same order of magnitude as the interannual variability.

@@

We estimated the biomass reductions in accordance to the CLs \citep{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature parameterization (\emph{subarctic}) for each bespoke PFT. The estimates are listed in Table~\ref{tab:biomass_reduction}. The reported range corresponds to the maximum deviation defined by the $\pm 20\,\unit{\%}$ variation of $f_\mathrm{light}$ with SWP=on. Total biomass of deciduous trees and perennial grassland was substentially reduced in comparison with the default MM parameterizations in both years. The CL for perennial grassland is higher than for the other species. From an agro-economical point of view in our focus area, where grass is cut only once a year, a loss of biomass of more than $10\,\unit{\%}$ can have severer consequences compared to more productive areas in central Europe and may need adjustment.

@@

@@

 The $\mathrm{DO_3SE}$ modeling results are in contrast to our observation of more visible damage in 2018 compared to 2019. However, visible damage caused by peak ozone uptake and a dependence of ozone sensibility on phenology are not accounted for in $\mathrm{POD_y}$. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland have been found to be more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed damage in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004} but not accounted for in the models. The literature results regarding ozone sensitivity of natural vegetation in northern Europe are contratictive. \citet{FS:Subramanian2014} reported a higher, modeled reduction of net primary production in conifers ($4.3-15.5\,\unit{\%}$) than birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, while \citet{Amb:Girgzdiene2009} observed more visible damage on deciduous trees than on Norway spruce in Lithuania. But visible damage and biomass reduction may not occur at the same rate and hence may not be interchangeable.

In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures than for central European species. To this end, we computed a PDF from observed temperature at Svanhovd (May--September, 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. Remark that these parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.

@@

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to a compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change strongly depend on both the range of existing acclimations as well as the actual type acclimation. This means vegetation which is acclimated to cold temperatures and little to heat- and drought stress might suffer more strongly from a more frequent occurance of heatwaves than plants which have a higher temperature tolerance.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on studies from less extreme climates. In particular, CLs for deciduous and coniferous trees were breached in all years and for all parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constrains on productivity leading to more severe economical consequences.

@@

\authorcontribution{All authors contributed to framing this research article, writing, and conclusions. SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV has performed the on-site observation of vegetation damage induced by ozone, provided advice on plant physiological processes, collected exisiting literature on subarctic vegetation, contributed significantly to framing and proof reading this research article. LE has contributed with expertize in $\mathrm{DO_3SE}$ modeling and development of the PDF-based temperature acclimation assessment. CO has collected PFT parameters from the literature, performed all $\mathrm{DO_3SE}$ simulations and validation. AE contributed with her experience regarding subarctic vegetation in Finnmark. FS contributed with an assessment of the 2018 meteorological conditions and writing the proposal for founding this project. TB gave valuable guidance in a broader research sense.} %% this section is mandatory

@@

\begin{acknowledgements}
  We thank NILU for performing the ozone measurements at Svanhovd for us, Bj{\o}rg Rognerud (Department of Geosciences, UiO) for processing SeNorge.no data with respect to beginning of growing season, Tore Flatlandsmo Berglen (NILU) for hourly pressure data from Svanhovd, NIBIO Environment Center Svanhovd for establishing and running the ozone garden both years, and Volkmar Timmermann (NIBIO) for sharing ICP Forest data from Svanhovd. This work was supported by the Research Council of Norway (Grant No. 268073).
\end{acknowledgements}
